Changes in accessibility to food banks and food services during COVID-19 and implications for the food security of vulnerable populations in Hamilton, Ontario

### 2021-04-24 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.Rmd
Add text to accessibility change interpretation

Travel time tables for three modes (car, transit, walking) were computed using the parcels as the origins and the locations of the food banks as the destinations. For routing, the package `r5r` [@pereira2021r5r] was used with a network extract for the City of Hamilton from OpenStreetMaps and the General Transit Feed Specification from Hamilton Street Railway, the local transit operator. For routing purposes we used maximum values of 180 min and 10,000 m walking distance: any destination that exceeded these thresholds was ignored. The departure time used for routing was **TIME**.

@@

Figure \ref{fig:sensitivity-analysis} shows the results of conducting sensitivity analysis of the system-wide accessibility as we vary the threshold (considering the situation before the pandemic). There is a clear pattern whereby more strict values of $\delta$ are associated with higher levels of system-wide accessibility: while increases in accessibility that result from decreases in the travel time window might seem counter-intuitive, this is a result of lower _congestion_, since fewer households are serviced and thus competition for the same resources is more limited. System-wide accessibility declines with higher values of $\delta$: as more households are serviced, congestion grows and the level of service declines, although this happens at a declining rate. We are not aware of any research that explain how long people are expected to travel for food banks, but we note that in developing countries, access to drinking water is defined as a source that takes less than 30 minutes to reach [round trip, see @world2019progress]. There is no reason why people in affluent countries should be expected to travel more for a basic need such as food. Accordingly, we adopt a 15-minute threshold for the analysis (representing a one-way trip). This threshold is also approximately where the rate of change in accessibility slows down.

@@

Turning to the location of individual facilities, the levels of service offered before and during the pandemic are shown in Figure \ref{fig:levels-of-service}. The level of service is functionally the inverse of the number of low-income households in the travel-mode weighted travel time catchment area of the facilities (this is because $S_j=1 \forall j$, i.e., each location represents a "capacity" of 1). Higher values mean that a facility is expected to service fewer households. Conversely, lower values indicate grater congestion.

The general pattern of the levels of service is similar before and during the pandemic, with lower values in the center of the city. Three more peripheral facilities towards the south of the city also have lower levels of service, presumably because they are expected to service relatively large suburban/exurban populations. During the pandemic, however, the levels of service dropped, in some cases quite substantially. The pattern of the losses in level of service, moreover, is not uniform. The upper pane of Figure \ref{fig:levels-of-service-changes} shows that the three peripheral facilities in the southern suburban/exurban part of the city had low levels of service to begin with, but did not see major declines during the pandemic. Further, the inset map shows that the levels of service deteriorated more in the central part of the city. However, the loss of level of service was not as large in the core (where most of the food banks/services are found), but instead was more marked in the inner ring around the core, where facilities may have faced greater demand from both central city and suburban populations after the closure of food banks during the pandemic.

@@

To further elucidate this issue, we now turn to the results of the accessibility analysis. As with the level of service of individual facilities, the general pattern of accessibility before and during the pandemic is similar. Figure \ref{fig:accessibility} reveals that, compared with the outer rural zones, the more urban zones of the city generally exhibit higher accessibility to food banks and food service locations. However, the pattern is not particularly smooth - this is largely attributable to the weighting of travel times by mode of transportation according to the trip patterns of low-income household respondents according to the TTS. For example, in zones where low-income households make a high proportion of trips by walking, access to food bank locations by walking is afforded a concomitantly high weight in our calculations of travel time compared to transit or car travel. From this, highly-accessible locations result from a mix of characteristics: low-income households in locations where travel options that align with zonal modal split are available to connect them to food bank locations with high levels of service within 15 minutes. This seems to track with the experience of some users of these services, as reported by Smith-Carrier et al. [-@smith2017food].

@@

We find that the accessibility landscape deteriorated substantially during the pandemic, with accessibility dropping on average by almost `r accessibility_taz_change %>% st_drop_geometry() %>% summarize(d = (.[[6]] - .[[5]])/.[[5]] * 100) %>% pull(d) %>% mean() %>% abs() %>% round()`\%, but with large variations: some zones experienced changes in accessibility of only about `r accessibility_taz_change %>% st_drop_geometry() %>% summarize(d = (.[[6]] - .[[5]])/.[[5]] * 100) %>% pull(d) %>% max() %>% abs() %>% round()`\%, whereas the most affected zone saw a loss of accessibility of almost `r accessibility_taz_change %>% st_drop_geometry() %>% summarize(d = (.[[6]] - .[[5]])/.[[5]] * 100) %>% pull(d) %>% min() %>% abs() %>% round()`\%. Figure \ref{fig:accessibility-changes-with-local-i} shows the changes in accessibility. Every zone is worse off after the closure of facilities during the pandemic, but some parts of the city seem to have been particularly affected. We used a local indicator of spatial autocorrelation [@anselin1995local] to explore the pattern of change in accessibility. Twenty-four TAZ are flagged as having significant large losses of accessibility (at $p<=0.10$, without correcting for multiple comparisons). Those zones are highlighted in the figure, where it can be seen that they form more or less compact neighborhoods. Remarkably, the largest (and significant) drops in accessibility are not downtown, but located in two cases in the industrial north of the city, in one case in an inner suburb above the escarpment (**need to describe the study area!**), and lastly in a more suburban/exurban region in the south-west (**share other insights about these regions, maybe based on weighted travel patterns?**).

@@

>>>>

Travel time tables for three modes (car, transit, walking) were computed using the parcels as the origins and the locations of the food banks as the destinations. For routing, the package `r5r` [@pereira2021r5r] was used with a network extract for the City of Hamilton from OpenStreetMaps and the General Transit Feed Specification from Hamilton Street Railway, the local transit operator. For routing purposes we used maximum values of 180 min and 10,000 m walking distance: any destination that exceeded these thresholds was ignored. The departure time used for routing was 8:00AM on March 30, 2021 to reflect transit service around the morning service peak on a typical Tuesday.

@@

Figure \ref{fig:sensitivity-analysis} shows the results of conducting sensitivity analysis of the system-wide accessibility as we vary the threshold (considering the situation before the pandemic). There is a clear pattern whereby more strict values of $\delta$ are associated with higher levels of system-wide accessibility: while increases in accessibility that result from decreases in the travel time window might seem counter-intuitive, this is a result of lower _congestion_, since fewer households are serviced and thus competition for the same resources is more limited. System-wide accessibility declines with higher values of $\delta$: as more households are serviced, congestion grows and the level of service declines, although this happens at a declining rate. We are not aware of any research that explains how long people are expected to travel for food banks, but we note that in developing countries, access to drinking water is defined as a source that takes less than 30 minutes to reach [round trip, see @world2019progress]. There is no reason why people in affluent countries should be expected to travel more for a basic need such as food. Accordingly, we adopt a 15-minute threshold for the analysis (representing a one-way trip). This threshold is also approximately where the rate of change in accessibility slows down.

@@

Turning to the location of individual facilities, the levels of service offered before and during the pandemic are shown in Figure \ref{fig:levels-of-service}. The level of service is functionally the inverse of the number of low-income households in the travel-mode weighted travel time catchment area of the facilities (this is because $S_j=1 \forall j$, i.e., each location represents a "capacity" of 1). Higher values mean that a facility is expected to service fewer households. Conversely, lower values indicate greater congestion.

The general pattern of the levels of service is similar before and during the pandemic, with lower values in the center of the city where low-income households exhibit multimodal trip patterns. Three more peripheral facilities towards the south of the city also have lower levels of service, presumably because they are expected to service relatively large suburban/exurban populations generally reliant on automobiles for travel. During the pandemic, however, the levels of service dropped, in some cases quite substantially. The pattern of the losses in level of service, moreover, is not uniform. The upper pane of Figure \ref{fig:levels-of-service-changes} shows that the three peripheral facilities in the southern suburban/exurban part of the city had low levels of service to begin with, but did not see major declines during the pandemic. Further, the inset map shows that the levels of service deteriorated more in the central part of the city. However, the loss of level of service was not as large in the core (where most of the food banks/services are found), but instead was more marked in the inner ring around the core, where facilities may have faced greater demand from both central city and suburban populations after the closure of food banks during the pandemic.

@@

To further elucidate this issue, we now turn to the results of the accessibility analysis. As with the level of service of individual facilities, the general pattern of accessibility before and during the pandemic is similar. Figure \ref{fig:accessibility} reveals that, compared with the outer rural zones, the more urban zones of the city generally exhibit higher accessibility to food banks and food service locations. However, the pattern is not particularly smooth - this is largely attributable to the weighting of travel times by mode of transportation according to the trip patterns of low-income household respondents captured by the TTS. For example, in zones where low-income households make a high proportion of trips by walking, access to food bank locations by walking is afforded a concomitantly high weight in our calculations of travel time compared to transit or car travel. From this, highly-accessible locations result from a mix of characteristics: low-income households in locations where travel options that align with zonal modal split are available to connect them to food bank locations with high levels of service within 15 minutes. This seems to track with the experience of some users of these services, as reported by Smith-Carrier et al. [-@smith2017food].

@@

We find that the accessibility landscape deteriorated substantially during the pandemic, with accessibility dropping on average by almost `r accessibility_taz_change %>% st_drop_geometry() %>% summarize(d = (.[[6]] - .[[5]])/.[[5]] * 100) %>% pull(d) %>% mean() %>% abs() %>% round()`\%, but with large variations: some zones experienced changes in accessibility of only about `r accessibility_taz_change %>% st_drop_geometry() %>% summarize(d = (.[[6]] - .[[5]])/.[[5]] * 100) %>% pull(d) %>% max() %>% abs() %>% round()`\%, whereas the most affected zone saw a loss of accessibility of almost `r accessibility_taz_change %>% st_drop_geometry() %>% summarize(d = (.[[6]] - .[[5]])/.[[5]] * 100) %>% pull(d) %>% min() %>% abs() %>% round()`\%. Figure \ref{fig:accessibility-changes-with-local-i} shows the changes in accessibility. Every zone is worse off after the closure of facilities during the pandemic, but some parts of the city seem to have been particularly affected. To better highlight these changes, we used a local indicator of spatial autocorrelation [@anselin1995local] to explore the pattern of change in accessibility. Twenty-four TAZ are flagged as having significant large losses of accessibility (at $p<=0.10$, without correcting for multiple comparisons). Those zones are highlighted in the figure, where it can be seen that they form more or less compact neighborhoods. Remarkably, the largest (and significant) drops in accessibility are not downtown, but located in two cases in the industrial north of the city, in one case in an inner suburb above the escarpment, and lastly in a more suburban/exurban region in the south-west (**share other insights about these regions, maybe based on weighted travel patterns?**).

@@

For the more suburban clusters of zones, the decrease in accessibility is derived from the closure of locations throughout the city reachable by car. In the cluster of central suburban zones for example, low-income households in the outer ring of zones that exhibit medium to high decreases in accessibility within this cluster appear to be largely auto-dependent in their tripmaking, which each each exhibiting between `r round(modes %>% st_drop_geometry() %>% filter(TAZUID %in% c(5143, 5135, 5121, 5110, 5102, 5092, 5063)) %>% summarize(min(p_car)) *100)`-`r round(modes %>% st_drop_geometry() %>% filter(TAZUID %in% c(5143, 5135, 5121, 5110, 5102, 5092, 5063)) %>% summarize(max(p_car)) *100)`\% of their modal split for car trips. This results in the parcels within these zones having a large number of potentially accessible locations in the travel time matrix. But by extension, the change in accessibility over the pre- and during-COVID-19 time periods is affected not only by the closure of service locations proximate to the zones, but also the locations in the central city. The zone with the greatest decrease in accessibility within this cluster (`r round(accessibility_taz_change %>% st_drop_geometry() %>% filter(TAZUID == 5135) %>% pull(acc_d), 4)`) has a high rate of car trips and connects to the most food bank locations in total as well as those that stayed open or closed. The exception to this is the zone in the middle of this cluster, which has a much higher proportion of transit use `r round(modes %>% st_drop_geometry() %>% filter(TAZUID == 5116) %>% summarize(first(p_transit)) *100)`\%. This modal split weighting results in fewer food bank and service locations accessible from the parcels within this zone which, by extension, limits their access to closed locations reachable within 15 minutes in our calculations.

In the cluster to the south-west, the decrease in accessibility is predominately driven by the closure of a high level-of-service Community Meals provider. However, like the more central suburban zones, low-income households within this cluster also between 90% to 100% auto-dependent in their tripmaking in the TTS. The story is similar for the zone located in the north-west that exhibits the greatest decrease in accessibility. Here, low-income households responding to the TTS conducted 100% of their trips by car and, as a result, parcels within this zone have access to the second-highest number of food bank and service locations within 15 minutes. However, this also means zonal accessibilities are greatly affected by the number of closures throughout the city. Finally, in the city's north end and north-east zones, low-income households exhibit a mixture of tripmaking behaviour in the TTS. Households in some zones take transit more often and one zone in particular has 100\% of its trips by walking while in others, car use makes up 100\% of trips. For these zones, the decrease in accessibility tends to be a product of the closure of several inner-city food bank and service locations.
